---
title: "Introduction to R Programming"
subtitle: "Part II"
author: |
  | Wenjie Wang
  | Department of Statistics, UConn
date: April 6, 2018
---


```{r setup, echo = FALSE}
options(width = 100)

### other simple utility functions
## create a tag with some handy defaults
a_new <- function(x, href = x, target = "_blank", ...)
{
    htmltools::tags$a(x, href = href, target = target, ...)
}
## paste with forward slash
paste4 <- function(...) paste(..., sep = "/")

### define some constants
rstudio_github <- "https://github.com/rstudio"
shiny_homepage <- "https://shiny.rstudio.com"
shiny_examples_github <- paste4(rstudio_github, "shiny-examples")
shiny_gallery <- paste4(shiny_homepage, "gallery")
shiny_tutorial <- paste4(shiny_homepage, "tutorial")
shiny_cheatsheet <- paste4(rstudio_github, "cheatsheets/blob/master/shiny.pdf")
```


## Getting Started

- the slides of part I:
  [wenjie-stat.me/2018-01-19-siam/][slides-part1]

- these slides: [wenjie-stat.me/2018-04-06-siam/][slides-part2]

- source code of these slides and all the examples:
  [https://github.com/wenjie2wang/2018-04-06-siam/][source-repo]

- prerequisites: [R][cran] and possibly the [RStudio IDE][rstudio-ide] have
  been installed.

- a list of R packages needed for following examples:

  ```{r pkgs-slides, eval = FALSE}
  install.packages(
      c("bookdown", "data.table", "dplyr", "DT", "dygraphs", "htmltools",
        "leaflet", "microbenchmark", "plotly", "R6", "readr", "shiny")
  )
  ```

```{r need-packages, echo = FALSE}
##' Check, Install and Attach Multiple R packages Specified
##'
##' The function first Checks whether the packages given were installed. Then
##' install them if they are not, then attach them to the search path.
##'
##' @usage need.packages(pkg)
##' @param pkg A character vector specifying the packages needed to reproduce
##'     this document.
##' @param ... Other arguments passed to function \code{\link[base]require}.
##' @return NULL invisibly.
##' @examples
##' need.pacakges(c("ggplot2", "geepack"))
need.packages <- function(pkg, ...)
{
    new.pkg <- pkg[! (pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, repos = "https://cloud.r-project.org")
    foo <- function(a, ...) suppressMessages(require(a, ...))
    sapply(pkg, foo, character.only = TRUE)
    invisible(NULL)
}
need.packages(
    c("bookdown", "data.table", "dplyr", "DT", "dygraphs", "htmltools",
      "leaflet", "microbenchmark", "plotly", "R6", "readr", "shiny")
)
```


## Outline

- reading and writing data

- data processing and manipulation

- reproducible reports with R Markdown

- interactive data visualization with R Shiny

- ~~object oriented systems in R~~


## Session information

```{r session-info}
sessionInfo()
```


## Reading and writing data

- basic functions:

```r
read.table(), read.csv(), readLines(), load(), readRDS(), ...
write.table(), write.csv(), writeLines(), save(), saveRDS(), ...
```

- can be slow when working with large data

- some useful packages:
  - for plain-text rectangular data (such as csv, tsv, and fwf):
    **utils**, **[readr](https://readr.tidyverse.org)**,
    **[data.table][data-dot-table]**, ...
  - for data stored in other formats: **foreign**, **haven**, **openxlsx**, ...


## Some tricks for efficiently reading large text files using `read.table()`

- specify `nrows`: the maximum number of rows to read in
  - e.g., we may determine the number of rows of `data.csv` by
    `get_nrows("data.csv")`, where `get_nrows()` is a simple function as
    follows:

```{r get-nrows-function}
get_nrows <- function(file) {
    as.integer(
        system2("wc", sprintf("-l %s | awk '{print $1}'", file), TRUE)
    )
}
```

- specify `colClasses`: column classes
  - e.g., only read in the first ten rows (`nrows = 10`) and decide on the
    appropriate classes

- see [this short post][toby-hocking-post] by Toby Hocking for a nice summary


## Using **readr** package

```r
read_table(), read_csv(), read_tsv(), read_fwf(), read_lines(), ...
write_table(), write_csv(), write_tsv(), write_fwf(), write_lines(), ...
```

**readr** providers a few features that make it more user-friendly than base R:

- more consistent arguments' naming (e.g., `col_names` vs. `header` and
  `col_types` vs. `colClasses`)
- leave strings as is by default, and automatically parse common date/time
  formats
- able to read compressed files (e.g., `.gz`, `.bz2`, `.xz`, or `.zip`)
  automatically
- ...


## Using **data.table** package

```r
fread(), fwrite()
```

- `fread()`/`fwrite()` is similar to `read.table()`/`write.table()` but much
  faster and more convenient.
- All controls such as `sep`, `colClasses` and `nrows` are automatically
  detected (or guessed) in `fread()`.


## Example data

- randomly generate a "not so small" csv file: `data.csv`

```{r simu-large-data, cache = TRUE}
options(stringsAsFactors = FALSE)
set.seed(613); n <- 1e6
dat <- data.frame(foo = rnorm(n),
                  bar = rpois(n, 5),
                  alpha = sample(letters, size = n, replace = TRUE),
                  beta = as.Date(rpois(n, 10), origin = "2018-04-06"),
                  gamma = gl(5, k = n / 5, labels = LETTERS[1:5]))
str(dat)
write.table(dat, file = "data.csv", sep = ",", row.names = FALSE)
sprintf("File size: %.0f MB", file.info("data.csv")$size / 1024 ^ 2)
```

---

## Examples of reading csv file

```{r simple-read-table, eval = FALSE}
## simply use `read.csv`
dat1 <- read.csv("data.csv")
## or `read.table`
dat2 <- read.table("data.csv", sep = ",", header = TRUE)
## or specify `nrows` and `colClasses`
dat3 <- read.table("data.csv", sep = ",", header = TRUE,
                   nrows = get_nrows("data.csv") - 1,
                   colClasses = c("numeric", "integer",
                                  rep("character", 3)))
## use readr::read_csv
dat4 <- readr::read_csv("data.csv")
dat5 <- readr::read_csv("data.csv", col_types = c("diccc"),
                        n_max = get_nrows("data.csv") - 1)
## use data.table::fread
dat6 <- data.table::fread("data.csv")
dat7 <- data.table::fread("data.csv", header = TRUE, sep = ",",
                          nrows = get_nrows("data.csv") - 1)
## exercise:
##   check the resulting column types of each `data.frame` object.
##   is there any difference? what are the classes of `dat4`,...,`dat7`?
```

---


```{r read-table-benchmark, cache = TRUE}
library(microbenchmark)
microbenchmark(
    read.csv = read.csv("data.csv"),
    read.table = read.table("data.csv", sep = ",", header = TRUE),
    read.table_tricks = read.table("data.csv", sep = ",", header = TRUE,
                                   nrows = get_nrows("data.csv") - 1L,
                                   colClasses = c("numeric", "integer",
                                                  rep("character", 3))),
    read_csv = readr::read_csv("data.csv", col_types = cols()),
    read_csv_tricks = readr::read_csv("data.csv", col_types = c("diccc"),
                                      n_max = get_nrows("data.csv") - 1),
    fread = data.table::fread("data.csv"),
    fread_tricks = data.table::fread("data.csv", header = TRUE, sep = ",",
                                     nrows = get_nrows("data.csv") - 1,
                                     colClasses = c("numeric", "integer",
                                                    rep("character", 3))),
    times = 30, unit = "relative")
```


## Data processing and manipulation

> - It is said that data scientists spent about 80% of time on the process of
    cleaning and preparing the data.

> - How to make data cleaning as efficient and effective as possible?

> - In fact, base R provides a variety of useful functions for data processing
    and manipulation.

> - two popular add-on packages: **[dplyr](http://dplyr.tidyverse.org)** and
    **[data.table][data-dot-table]**


## dplyr: A grammar of data manipulation

some of the key "verbs":

- `select()`: picks columns/variables based on names or indices
- `filter()`: extracts a subset of rows based on logical conditions
- `arrange()`: reorders rows
- `rename()`: renames columns/variables
- `mutate()`: creates new columns/variables
- `summarise()` or `summarize()`: computes summary statistics
- `group_by()`: helps perform operations by group.


## Common properties

In particular,

- We may directly refer to columns without using `$` operator.
- The first argument is a data frame and the returned result is a new data frame.
- must be tidy data (Wickham, 2014): one observation per row, and each column
  representing a feature or characteristic of that observation


## `select()`

```{r dplyr-examples-select-1, cache = TRUE}
names(dat)
str(select(dat, foo, alpha))
str(select(dat, bar:beta))
str(select(dat, -(bar:beta)))
```

---

- `select()` helpers:

```r
starts_with(), ends_with(), contains(), matches(),
num_range(), one_of(), everything()
```

```{r dplyr-examples-select-2, cache = TRUE}
str(select(dat, starts_with("b")))
str(select(dat, ends_with("a")))
str(select(dat, -matches("^b|a$")))
```

---

## `filter()`

```{r dplyr-examples-filter-1, cache = TRUE}
str(filter(dat, foo > 2 & bar %in% c(3, 4)))
## str(filter(dat, foo > 2, bar %in% c(3, 4))) # equivalent
```

- hmmm... why inventing another wheel?

```{r dplyr-examples-filter-2, eval = FALSE}
## why not use `[.data.frame`?
subDat <- dat[dat$foo > 2 & dat$bar %in% c(3, 4), ]
## not a fan of `$` operator? okay! no `$` if using `with`, right?
subDat <- with(dat, dat[foo > 2 & bar %in% c(3, 4), ])
## how about using `base::subset`?
subDat <- base::subset(dat, foo > 2 & bar %in% c(3, 4))
```

## A quick benchmarking

```{r benchmark-subset, cache = TRUE}
dat_tbl <- tbl_df(dat)
dat_dt <- as.data.table(dat)
microbenchmark(
    "[_$" = dat[dat$foo > 2 & dat$bar %in% c(3, 4), ],
    "[_with" = with(dat, dat[foo > 2 & bar %in% c(3, 4), ]),
    "subset" = base::subset(dat, foo > 2 & bar %in% c(3, 4)),
    "filter_&" = dplyr::filter(dat, foo > 2 & bar %in% c(3, 4)),
    "filter_," = dplyr::filter(dat, foo > 2, bar %in% c(3, 4)),
    "filter_tbl" = dplyr::filter(dat_tbl, foo > 2, bar %in% c(3, 4)),
    "data.table" = dat_dt[foo > 2 & bar %in% c(3, 4)],
    times = 200, unit = "relative"
)
```

## `arrange()`

```{r dplyr-examples-arrange, cache = TRUE}
asc_dat <- arrange(dat, beta, foo)
head(select(asc_dat, beta, foo), 3)
tail(select(asc_dat, beta, foo), 3)
desc_dat <- arrange(dat, desc(beta), foo)
head(select(desc_dat, beta, foo), 3)
```

- again, why not use `base::order`?


## A quick benchmarking

```{r benchmark-order, cache = TRUE}
microbenchmark(
    "order_$" = dat[order(as.numeric(dat$beta), dat$foo,
                          decreasing = c(TRUE, FALSE)), ],
    "order_with" = with(dat, dat[order(as.numeric(beta), foo,
                                       decreasing = c(TRUE, FALSE)), ]),
    "arrange" = arrange(dat, desc(beta), foo),
    "arrange_as" = arrange(dat, desc(as.numeric(beta)), foo),
    "data.table" = dat_dt[order(- beta, foo), ],
    times = 100, unit = "relative"
)
```

## `rename()`

```{r dplyr-examples-rename-1, cache = TRUE}
names(rename(dat, x = foo, y = bar))
```

- how to do it with base R?

```{r dplyr-examples-rename-2, eval = FALSE}
names(dat)[names(dat) == "foo"] <- "x"
names(dat)[names(dat) == "bar"] <- "y"
```

- what about **data.table**?

```{r data-table-examples-setnames, eval = FALSE}
setnames(dat_dt, c("foo", "bar"), c("x", "y"))
```

## `mutate()`

```{r dplyr-examples-mutate-1, cache = TRUE}
str(mutate(dat, bar_centered = bar - mean(bar)))
```

- how to do it with base R?

```{r dplyr-examples-mutate-2, eval = FALSE}
dat$bar_centered <- dat$bar - mean(dat$bar)
## or using `with`
dat$bar_centered <- with(dat, bar - mean(bar))
```

- what about **data.table**?

```{r data-table-examples-newCol, eval = FALSE}
dat_dt[, `:=`(bar_centered = bar - mean(bar))]
```


## `summarize` and `group_by()`

```{r dplyr-examples-group-by, cache = TRUE}
dat_grouped <- group_by(dat, gamma)
summarise(dat_grouped, mean_foo = sd(foo), median_bar = mean(bar))
```

---

## Forward-pipe operator: `%>%`

- provided by **[magrittr](https://CRAN.R-project.org/package=magrittr)**
  package
- similar idea to the "piping" (using `|`) in Linux and other Unix-like
  operating systems
- examples of basic chaining or piping:
  - `x %>% f` $\Leftrightarrow$ `f(x)`
  - `x %>% f(y)` $\Leftrightarrow$ `f(x, y)`
  - `x %>% f %>% g %>% h` $\Leftrightarrow$ `h(g(f(x)))`
- example of argument placeholder
  - `x %>% f(y, .)` $\Leftrightarrow$ `f(y, x)`
- pros: clearly expressing a sequence of multiple operations
- cons: possibly hard to debug without intermediate steps


## Quick benchmarkings

- any performance compromised?

```{r benchmark-pipe-1, cache = TRUE}
microbenchmark(
    nested = summary(head(dat, n = 100)),
    steps = {
        tmpDat <- head(dat, n = 100)
        summary(tmpDat)
    },
    pipe = dat %>% head(n = 100) %>% summary,
    times = 1e3, unit = "relative"
)
```

---

```{r benchmark-pipe-2, cache = TRUE}
microbenchmark(
    nested = prop.table(xtabs(~ bar, head(dat, n = 100))),
    steps = {
        tmpDat <- head(dat, n = 100)
        tmpTab <- xtabs(~ bar, tmpDat)
        prop.table(tmpTab)
    },
    pipe = dat %>% head(n = 100) %>% xtabs(~ bar, .) %>% prop.table,
    times = 1e3, unit = "relative"
)
```

## Using **dplyr** with piping

```{r dplyr-example-all, cache = TRUE}
dat %>%
    select(- alpha) %>%
    rename(group = gamma) %>%
    mutate(abs_foo = abs(foo)) %>%
    group_by(group) %>%
    arrange(beta, desc(bar)) %>%
    head(n = 200) %>%
    summarize(median_abs_foo = median(abs_foo))
```


## Reproducible reports with R Markdown

- code + narratives = report
- some existing tools:
  - WEB (Donald Knuth, Literate Programming)
  - Noweb (Norman Ramsey)
  - Sweave (Friedrich Leisch and R-core)
  - **knitr** (Yihui Xie)
  - Org mode + Babel (Carsten Dominik, Eric Schulte, ...)
  - Jupyter notebook
- **knitr**:
  - `.Rnw` (R + LaTeX)
  - `.Rmd` (R + Markdown)
  - any computing language + any authoring language

---

- R Markdown (**rmarkdown**): http://rmarkdown.rstudio.com
  - [gallery][rmarkdown-gallery]
  - [cheatsheet][rmarkdown-cheatsheet]
- **bookdown**: https://bookdown.org
- examples:
  - these slides
  - a [project template][dslab-templates] written in R Markdown and **bookdown**
    for [Data Science Lab at UConn](http://statds.org)
  - learn from the other cool kids! http://yihui.name/knitr/demo/showcase/

---


## Interactive data visualization with R Shiny

- What is R Shiny?

- **shiny** is an R package that makes it easy to build interactive web
    applications straight from R.
  - homepage: `r a_new(shiny_homepage)`
  - a collection of toy examples by RStudio:
    `r a_new(shiny_examples_github)`
  - official `r a_new('gallery', shiny_gallery)`,
    `r a_new('tutorial', shiny_tutorial)`, and
    `r a_new('cheatsheet', shiny_cheatsheet)`
  - other galleries:
    `r a_new("Show me Shiny", "https://www.showmeshiny.com")`,
    `r a_new("Plotly", "https://plot.ly/r/shiny-gallery/")`, ...


---

## Some examples

- `r a_new("Hello Shiny!", "https://gallery.shinyapps.io/001-hello/")`
- `r a_new("Exploring Historical English Soccer Data",
           "https://www.showmeshiny.com/historical-english-soccer/")`
- `r a_new("Weekly influenza surveillance data",
           "https://github.com/kb230557/Flu_Shiny_App")`
- `r a_new("Paris car accidents",
           "http://www.remyzum.com/shiny/Paris_Accidentologie/")`
- `r a_new("Real estate analytics", "https://www.showmeshiny.com/shinyhome/")`
  using Shiny dashboard


## Structure of a Shiny app

By execution order:

1. `global.R`: an optional script for code needed in `ui.R` and
   `server.R`

    ```r
    library(shiny)              # e.g., attach the Shiny package
    ```

2. `ui.R`: define user interface (UI) design

    ```r
    shinyUI(fluidPage(
        ## e.g., create a fluid page
    ))
    ```

3. `server.R`: define server-side logic

    ```r
    shinyServer(function(input, output, session) {
        ## a function with two required argument, `input` and `output`,
        ## and an optional argument `session`
    })
    ```


## Alternative structure

- a sinlge script called `app.R`

```r
library(shiny)

## e.g., create a fluid page
ui <- fluidPage(
    ## some ui elements and widgets
)

## define server-side logic
server <- function(input, output, session) {
    ## the server function body
}

shinyApp(ui = ui, server = server)
```


## Sidebar layout

```{r sidebar-layout}
ui <- fluidPage(
    titlePanel("my title panel"),
    sidebarLayout(
        sidebarPanel("my sidebar panel"),
        mainPanel("my main panel")
    )
)
```

<hr/>

```{r sidebar-layout-preview, echo = FALSE, results = "asis"}
ui
```

---

- Shiny uses `r a_new("Twitter Bootstrap 3",
  "https://getbootstrap.com/docs/3.3/")`, the probably most popular HTML, CSS,
  and JS framework for developing responsive, mobile first projects on the web.

- the essential HTML code defined in `ui`:

  ```{r sidebar-layout-html, echo = FALSE, comment = "", class.output = "sourceCode html"}
  print(ui)
  ```

- The more fundamental layout constructor functions are `fluidRow()` and
  `column()`.


## HTML builders

<!--
    the following table source is mainly taken from
    http://shiny.rstudio.com/tutorial/written-tutorial/lesson2/
-->

<table>
  <thead>
    <tr>
      <th>Shiny</th>
      <th>HTML5</th>
      <th>creates</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td><code>p()</code></td>
      <td><code>&lt;p&gt;</code></td>
      <td>A paragraph of text</td>
    </tr>
    <tr>
      <td><code>h1(), ..., h6()</code></td>
      <td><code>&lt;h1&gt;, ..., &lt;h6&gt;</code></td>
      <td>a first, ..., sixth level header</td>
    </tr>
    <tr>
      <td><code>a()</code></td>
      <td><code>&lt;a&gt;</code></td>
      <td>A hyper link</td>
    </tr>
    <tr>
      <td><code>br()</code></td>
      <td><code>&lt;br&gt;</code></td>
      <td>A line break</td>
    </tr>
    <tr>
      <td><code>div()</code></td>
      <td><code>&lt;div&gt;</code></td>
      <td>A division with a uniform style</td>
    </tr>
    <tr>
      <td><code>span()</code></td>
      <td><code>&lt;span&gt;</code></td>
      <td>An in-line version of division</td>
    </tr>
    <tr>
      <td><code>strong()</code></td>
      <td><code>&lt;strong&gt;</code></td>
      <td>Bold text</td>
    </tr>
    <tr>
      <td><code>em()</code></td>
      <td><code>&lt;em&gt;</code></td>
      <td>Italicized text</td>
    </tr>
    <tr>
      <td><code>...</code></td>
      <td><code>...</code></td>
      <td>...</td>
    </tr>
    <tr>
      <td><code>HTML()</code></td>
      <td>&nbsp;</td>
      <td>Directly passes character strings as HTML</td>
    </tr>
  </tbody>
</table>


- **shiny** imports HTML builder functions from **htmltools**.
- `names(tags)` returns a complete valid HTML5 tag list.


## Basic widgets

- What is a web widget? A web element that users can interact with.
- `r a_new("Standard widgets gallery",
  "https://shiny.rstudio.com/gallery/widget-gallery.html")`
- The first two arguments for each widget function are
  - `id` for widget name: users will not see the name, but you can use it
    to access the widget's value. The name should be a character string.
  - `label` for widget label: this label will appear with the widget in your
    app. It should be a character string, but it can be an empty string `""`.

---

The standard Shiny widgets include

<!--
    table source directly taken from
    http://shiny.rstudio.com/tutorial/written-tutorial/lesson3/
-->

<table>
  <thead>
    <tr>
      <th>function</th>
      <th>widget</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td><code>actionButton</code></td>
      <td>Action Button</td>
    </tr>
    <tr>
      <td><code>checkboxGroupInput</code></td>
      <td>A group of check boxes</td>
    </tr>
    <tr>
      <td><code>checkboxInput</code></td>
      <td>A single check box</td>
    </tr>
    <tr>
      <td><code>dateInput</code></td>
      <td>A calendar to aid date selection</td>
    </tr>
    <tr>
      <td><code>dateRangeInput</code></td>
      <td>A pair of calendars for selecting a date range</td>
    </tr>
    <tr>
      <td><code>fileInput</code></td>
      <td>A file upload control wizard</td>
    </tr>
    <tr>
      <td><code>helpText</code></td>
      <td>Help text that can be added to an input form</td>
    </tr>
    <tr>
      <td><code>numericInput</code></td>
      <td>A field to enter numbers</td>
    </tr>
    <tr>
      <td><code>radioButtons</code></td>
      <td>A set of radio buttons</td>
    </tr>
    <tr>
      <td><code>selectInput</code></td>
      <td>A box with choices to select from</td>
    </tr>
    <tr>
      <td><code>sliderInput</code></td>
      <td>A slider bar</td>
    </tr>
    <tr>
      <td><code>submitButton</code></td>
      <td>A submit button</td>
    </tr>
    <tr>
      <td><code>textInput</code></td>
      <td>A field to enter text</td>
    </tr>
  </tbody>
</table>


## Reactive output

`*Output` functions in `ui` or `ui.R` turn R objects into output of UI.

<!--
    table source taken from
    http://shiny.rstudio.com/tutorial/written-tutorial/lesson4/
-->

<table>
  <thead>
    <tr>
      <th>Output function</th>
      <th>Creates</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td><code>htmlOutput</code></td>
      <td>raw HTML</td>
    </tr>
    <tr>
      <td><code>imageOutput</code></td>
      <td>image</td>
    </tr>
    <tr>
      <td><code>plotOutput</code></td>
      <td>plot</td>
    </tr>
    <tr>
      <td><code>tableOutput</code></td>
      <td>table</td>
    </tr>
    <tr>
      <td><code>textOutput</code></td>
      <td>text</td>
    </tr>
    <tr>
      <td><code>uiOutput</code></td>
      <td>raw HTML</td>
    </tr>
    <tr>
      <td><code>verbatimTextOutput</code></td>
      <td>text</td>
    </tr>
  </tbody>
</table>

- These `*Output` functions take output name/ID as input.

---

`render*` functions in `server` or `server.R`

<!--
    table source taken from
    http://shiny.rstudio.com/tutorial/written-tutorial/lesson4/
-->

<table>
  <thead>
    <tr>
      <th>render function</th>
      <th>creates</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <td><code>renderImage</code></td>
      <td>images (saved as a link to a source file)</td>
    </tr>
    <tr>
      <td><code>renderPlot</code></td>
      <td>plots</td>
    </tr>
    <tr>
      <td><code>renderPrint</code></td>
      <td>any printed output</td>
    </tr>
    <tr>
      <td><code>renderTable</code></td>
      <td>data frame, matrix, other table like structures</td>
    </tr>
    <tr>
      <td><code>renderText</code></td>
      <td>character strings</td>
    </tr>
    <tr>
      <td><code>renderUI</code></td>
      <td>a Shiny tag object or HTML</td>
    </tr>
  </tbody>
</table>

- These `render*` functions take a single argument: an R expression surrounded
  by `{}`.
- Shiny runs the R expressions inside `render*` functions once each time a user
  changes the value of a widget.

---

<div class="full-height">
```r
ui <- fluidPage(
    titlePanel("a simple example of reactive output"),
    sidebarLayout(
        sidebarPanel(
            selectInput("cyl", "Cylinders",
                        choices = sort(unique(mtcars$cyl)),
                        selected = 4)
        ),
        mainPanel(plotOutput("mpg_boxplot"))
    )
)

server <- function(input, output) {
    ## the `renderPlot` runs every time a user changes input$cyl
    output$mpg_boxplot <- renderPlot({
        ## filtering the cars models in the mtcars dataset
        ## by the input cylinder number
        dat <- subset(mtcars, cyl == input$cyl)
        ## draw a boxplot of mpg for the filtered data
        with(dat, boxplot(mpg))
    })
}

shinyApp(ui, server)
```
</div>

---

```{r mtcars-app, echo = FALSE}
knitr::include_app("https://shiny.wenjie-stat.me/examples/mtcars/",
                   height = "600px")
```


## Reactive expressions

- Create a reactive expression by the `reactive()` function, which takes
  an R expression surrounded by `{}` similar to `render*` functions.

```r
server <- function(input, output, session) {
    ## e.g., a simple reactive expression for filtering the cars models
    ## in the mtcars dataset by the input cylinder
    dataInput <- reactive({
        subset(mtcars, cyl == input$cyl)
    })
    ## draw a boxplot of mpg for the filtered data
    output$mpg_boxplot <- renderPlot({
        dat <- dataInput()
        with(dat, boxplot(mpg))
    })
}
```

- Reactive expressions cache values and update them only when it is necessary,
  which make the app faster.


## Share Shiny apps

- share as R scripts by `runApp()`, `runUrl()`, `runGitHub()` or `runGist()`

- share as a web page:
  - `r a_new("https://www.shinyapps.io")`
    - free for 5 apps and 25 active hours per month
  - Shiny server open source
    - very basic features
    - e.g., my Shiny server: `r a_new("https://shiny.wenjie-stat.me")`
  - Shiny server Pro
  - RStudio Connect


## More advanced HTML widgets

- **`r a_new("htmlwidgets", "http://www.htmlwidgets.org/")`**
  package provides a framework that helps to bring the best of
  JavaScript data visualization to R.
- example R packages built with **htmltwidgets**:
  - **`r a_new("leaflet", paste4(rstudio_github, "leaflet"))`**
    for geo-spatial mapping powered by JavaScript library leaflet
  - **`r a_new("dygraphs", paste4(rstudio_github, "dygraphs"))`**
    for time series charting powered by JavaScript library dygraphs
  - **`r a_new("plotly", "https://plot.ly/r/")`** and
    **`r a_new("Highcharter", "http://jkunst.com/highcharter/")`**
    for general interactive graphics powered by
    JavaScript library plotly.js and Highcharts, respectively.
  - **`r a_new("DT", paste4(rstudio_github, "DT"))`** for tabular data display
    powered by JavaScript library DataTables.


## leaflet example

- locations of earthquakes off Fiji

```{r leaflet-app, echo = FALSE}
knitr::include_app("https://shiny.wenjie-stat.me/examples/leaflet/",
                   height = "500px")
```


## dygraphs example

```{r dygraphs-app, echo = FALSE}
knitr::include_app("https://shiny.wenjie-stat.me/examples/dygraphs/",
                   height = "500px")
```


## plotly example

```{r plotly-app, echo = FALSE}
knitr::include_app("https://shiny.wenjie-stat.me/examples/plotly/",
                   height = "600px")
```


## DT example

```{r dt-app, echo = FALSE}
knitr::include_app("https://shiny.wenjie-stat.me/examples/DT/",
                   height = "600px")
```


## More options on UI and control widgets

- **`r a_new("shinythemes", paste4(rstudio_github, "shinythemes"))`**
  includes several Bootstrap themes from
  `r a_new("Bootswatch", "https://bootswatch.com/")`.
- **`r a_new("shinydashboard", paste4(rstudio_github, "shinythemes"))`**
  makes it easy to use Shiny to create dashboards.
- **`r a_new("miniUI", paste4(rstudio_github, "miniUI"))`** provides UI widget
  and layout functions for writing Shiny apps that work well on small screens.
- **`r a_new("shinyWidgets", "https://github.com/dreamRs/shinyWidgets")`**
  provides more control widgets that are not available in **shiny** package.
- **`r a_new("shinyAce", "https://github.com/trestletech/shinyAce")`**
  integrates the `r a_new("Ace text editor", "https://ace.c9.io")` with Shiny.
- **`r a_new("shiny.semantic", "https://github.com/Appsilon/shiny.semantic")`**
  provides a wrapper for `r a_new("Semantic UI", "https://semantic-ui.com")`.
- ...



## Some reference and further reading

- [*R Programming for Data
  Science*](https://bookdown.org/rdpeng/rprogdatascience/) by Roger D. Peng
- [*R for Data Science*](http://r4ds.had.co.nz/) by Garrett Grolemund, and
  Hadley Wickham
- [*Advanced R*](http://adv-r.had.co.nz) by Hadley Wickham
- [*Dynamic documents with R and knitr*](https://github.com/yihui/knitr-book) by
  Yihui Xie
- Wickham, H. (2014). Tidy Data. Journal of Statistical Software, 59(10), 1-23.
- `r a_new("a collection of Shiny tips & tricks",
  "https://github.com/daattali/advanced-shiny")`
- `r a_new("modularizing Shiny app code",
  "https://shiny.rstudio.com/articles/modules.html")`
- **`r a_new("shinytest", paste4(rstudio_github, "shinytest"))`** for automated
  testing for Shiny apps
- `r a_new("more articles", "https://shiny.rstudio.com/articles/")` from RStudio


## Thanks and happy coding!

- https://wenjie-stat.me
- https://github.com/wenjie2wang/
- wenjie.2.wang@uconn.edu

[slides-part1]: https://wenjie-stat.me/2018-01-19-siam/
[slides-part2]: https://wenjie-stat.me/2018-04-06-siam/
[source-repo]: https://github.com/wenjie2wang/2018-04-06-siam/
[cran]: https://cran.r-project.org
[rstudio-ide]: https://www.rstudio.com/products/RStudio/#Desktop
[toby-hocking-post]:
    http://members.cbio.mines-paristech.fr/~thocking/reading-large-text-files-into-R.html
[data-dot-table]: https://github.com/Rdatatable/data.table/wiki
[rmarkdown-cheatsheet]:
    https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf
[rmarkdown-gallery]: http://rmarkdown.rstudio.com/gallery.html
[dslab-templates]: https://github.com/statds/dslab-templates
[shiny-gallery]: http://shiny.rstudio.com/gallery/
[shiny-cheatsheet]: http://shiny.rstudio.com/images/shiny-cheatsheet.pdf

